---
title: "Airline Data Exploration"
author: "Cofucan"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This is an EDA of flight data from 1987 to 2008. It is a very large dataset which has been compressed into csv files for each year. For my analysis, I will only be working with data from 2004 to 2008

## Load Libraries

```{r}
library(tidyverse)
library(skimr)
library(here)
library(janitor)
```

#### Reading in the column descriptions csv

```{r}
metadata <- read.csv("airline_dataset_column_info.csv")
view(metadata)
```

#### Reading in the data

```{r}
# 3004 data
df_airline <- read.csv("O:/GitHub/data_analysis/datasets/airline/2004.csv")
as_tibble(df_airline)
glimpse(df_airline)
```

#### Data Summary & Stats

```{r}
skim_df = skim_without_charts(df_airline)
# skim(df_airline)
skim_df
```

#### Sort by arrival time in descending order

```{r}
sorted_arr = df_airline %>% arrange(-ArrTime)
sorted_arr
```

#### Average arrival time for each month of 2004

```{r}
mean_arr <- df_airline %>% group_by(Month) %>% drop_na() %>% summarise(MeanArrivalPerMonth = mean(ArrTime))
mean_arr
```

## Python Test

```{python}
import pandas as pd
import numpy as np

hotels = pd.read_csv("C:/Users/Cofucan/Desktop/hotel_bookings.csv")
hotels.info()

print("hello")
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```
